\documentclass[12pt,letterpaper,cm]{hmcpset}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm2e}
\usepackage{enumerate,hyperref}

% info for header block in upper right hand corner
\name{-----------}
\class{Math 173}
\assignment{Problem Set 4}
\duedate{October 8, 2018}
\setlength\parindent{0pt}

\begin{document}

\begin{problem}[1]
    Give an example of two norms on a vector space $\mathcal{V}$ that are \emph{not} equivalent.
\end{problem}

\begin{solution}
    \vfill
\end{solution}

\begin{problem}[2]
    Let $(\mathcal{V},\|\cdot\|)$ be a normed space. Prove that
    the closed unit ball $B = \{x\in \mathcal{V} : \|x\|\leq 1\}$ is compact if and
    only if $\mathcal{V}$ is finite dimensional. \textit{Hint:} Find a sequence in
    $B$ that has no convergent subsequence; use norm equivalence.
\end{problem}

\begin{solution}
    \vfill
\end{solution}

\begin{problem}[3]
    Let $A\in\mathsf{M}_n(\mathbb{R})$. Prove that the Schatten-$p$ norm
    $\|A\|_p^p = \operatorname{tr}(|A|^p)$, where the operator absolute value
    $|A|$ is equal to the positive semidefinite square root of $A^*A$.
\end{problem}

\begin{solution}
    \vfill
\end{solution}

\begin{problem}[4]
    Let $A\in\mathbb{C}^{m\times n}$ and decompose $A = U \Sigma V^*$ into its singular value
    decomposition where $\Sigma$ is the diagonal matrix composed of the singular values
    $\sigma_1\geq\sigma_2\geq \dots\geq 0$. Write $\Sigma_k = \operatorname{diag}\{\sigma_1,\sigma_2
    ,\ldots,\sigma_k,0,0,\ldots\}$ for $k=1,2,\ldots,\max\{m,n\}$ and define the truncated singular value decomposition
    $A_k = U\Sigma_k V^*$. Prove\footnote{As we'll see later in the course, this says that the distance between
    $A$ and $A_k$ in the so-called \emph{spectral norm} is $\sigma_{k+1}$. It turns out that this is the
    \emph{best} rank-$k$ approximation to $A$ in both the spectral and Frobenius norms.}
    that the largest singular value of $A - A_k$ is $\sigma_{k+1}$.
\end{problem}

\begin{solution}
    \vfill
\end{solution}

\begin{problem}[5]
    Load the image located at \url{https://math173.github.io/img/dog.jpg} into a computational language of your choice.
    We can think of this grayscale image as a matrix $A\in\mathbb{R}^{m\times n}$. Where each element $A_{ij}$ is
    the intensity of the pixel $A_{ij}$.
\begin{enumerate}[(a)]
    \item Plot\footnote{You can use an SVD function. No need to implement it.} (in decreasing order) the singular values of $A$.
    \item Plot the truncated SVD $A_k$ as an image for $k=5,10,15,50$ and compare it to $A$.
    \item How many entries do we need to store to represent $A_k$ for general $A\in\mathbb{R}^{m\times n}$?
        Using this, interpret $A_k$ as a lossy compression scheme for a matrix $A$.
\end{enumerate}
\end{problem}

\begin{solution}
    \vfill
\end{solution}
\clearpage

\begin{problem}[Extra (Not Optional)]
    Write a short (less than half a page, more than a quarter) but thoughtful proposal
    for your midterm project. Be concrete about what specific problem in linear algebra
    (broadly defined) you'd like to investigate.
    Consult with the teaching staff if you'd like any assistance finding a topic.
\end{problem}

\begin{solution}
    \vfill
\end{solution}
\end{document}
